Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2019-07-08 GWLDCI.tex
moved the introduction around a bit

This paper is the first in a two part series investigating the combinatorics and geometry underlying SYM $N=4$ theory. This series lays out several results about the relationship between Wilson loop diagrams and the positroid cells that give a CW complex structure to the positive Grassmannian. This paper expands on an observation from \cite{wilsonloop}, that shows that certain families of Wilson loop diagrams correspond to the same positroid cell. This paper gives a way of enumerating the number of different Wilson loop diagrams that correspond to a given positroid cell. Furthermore, it shows that the number of cells in the image of the map to positive Grassmannians is given by counting the directions \hlfix{in}{of a?} associahedra.

The second paper in this series identifies which positroid cell is associated to which Wilson loop diagram \cite{generalcombinatoricsI}. It proves that a Wilson loop diagram with $k$ propagators corresponds to a $3k$ dimensional positroid cell. Furthermore, it gives an algorithm for passing from a Wilson loop diagram to the Grassmann necklace that identifies the associated positroid cell. In this manner the authors give a means of identifying which positroid cell is associated to which Wilson loop diagram that circumvents the painstaking process of examining bases sets. Finally, the authors characterize the the volume forms that Wilson loop diagrams associate to each positroid cells.

In recent years, there has been an active program researching the geometry and combinatorics underlying SYM $N=4$ theory \cite{wilsonloop, Arkani-Hamed:2013jha, Amplituhedronsquared, AmplituhedronDecomposition, galashinlam}. This body of work started with the observation that the on shell (tree level) amplitudes of this theory correspond to the volume of a subspace of a positive Grassmannian, called an Amplituhedron \cite{Arkani-Hamed:2013jha}. Since then, there has been significant work studying the structure of the amplituhedron, geometrically and combinatorially, for instance \cite{Arkani-Hamed:2013kca, arkani:2012nw, AmplituhedronDecomposition, galashinlam}. The focus of much of this work has been on understanding the relationship between the combinatorial objects from the physics of the system, BCFW diagrams, and their corresponding natural object for studying the positive Grassmannians, positroid cells.

Meanwhile, there has been a different body of work studying SYM $N=4$ theory from the point of view of Feynman integrals in twistor space \cite{Adamo:2011pv, Britto:2005fq, Cachazo:2004kj}. These integrals are calculated via Wilson loops. The \hlfix{first author and Amat}{At other points, we say "some of us" instead of specifying like this. Choose a convention} uncovered a connection between the Feynman integrals developed in this literature and positroids, defined  as matroids that are realized as an element of $\Gr(k,n)$ \cite{wilsonloop}. With the second author and others, this work was extended to study these Feynman integrals in terms of positroid cells and the positive Grassmannians \cite{casestudy, non-orientable}. Others have tried different approaches to define a geometric space associated to these integrals in a manner similar to the Amplituhedron \cite{Amplituhedronsquared, HeslopStewart}.

Both the Amplituhedron literature and the Wilson loop literature associate $n$ particle, $N^kMHV$ interaction in SYM $N=4$ theory to volume forms on a set of positroid cells of $\Gr(k,n)$. This gives a CW complex of some submanifold in $\Gr(k,n)$. Understanding the geometry and topology of these spaces is the subject of ongoing research. For Wilson loop diagrams with $k=2$ and $n=6$, past work of some of us shows that the manifold is not contractible \cite{casestudy}. Furthermore, the spaces defined by the Amplituhedron and the Wilson loop diagrams are not the same \cite{Amplituhedronsquared}. In particular, it it more natural to consider Deodhar cells of $\mathbb{G}(k, n+1)$ for Wilson loop diagrams. In this case, one gets that the manifold represented by the diagrams is not orientable \cite{non-orientable}.

In this paper, we focus on the correspondence between Wilson loop diagrams and positroids, first defined in \cite{wilsonloop}. In that paper, \hlfix{one of us}{see previous comment about self reference convention} with Amat show that there are certain families of diagrams, specifically those with exact subdiagrams, that do map to the same positroid cell. In \cite{casestudy} some of us explicitly show that not every positroid cell of a given dimension corresponds to a Wilson loop diagram. In this paper, we better enumerate this correspondence. We show that the map from Wilson loop diagrams to positroid cells is one to one exactly on the subset of diagrams with no non-trivial exact subdiagrams. We show that exact diagrams correspond to uniform matroids, and that contraction by exact subdiagrams has special matroidal properties. Further, given a Wilson loop diagram with a non-trivial exact subdiagram, we count the total number of Wilson loop diagrams that correspond to that same positroid cell. While we do not count the number of distinct positroid cells that correspond to Wilson loop diagrams, we do show that this enumeration is equivalent to enumerating the number of directions of an associahedron.

@@

We proceed in Section~\ref{section overall background} with the required definitions for Wilson loop diagrams (subsection~\ref{section WLD background}) and matroids (subsection~\ref{sec matroid background}).  Wilson loop diagrams  are said to be admissible when they correspond to $N^kMHV$ diagrams, or mathematically according to Definition~\ref{admisdfn}. The other key definition in this section is Definition~\ref{equivdfn}, which gives an equivalence relation on admissible Wilson loop diagrams.  The connection to matroids is stated precisely in Theorem \ref{thm WLD defines matroid}.

The main result of Section~\ref{sec equiv} is Theorem~\ref{same matroid iff equiv}, which says that two Wilson loop diagrams define the same matroid if and only if they are equivalent.  In subsection~\ref{sec: polygon partitions} we relate certain subdiagrams (the {\em exact} subdiagrams, Definition~\ref{def:exact diagram}) of Wilson loop diagrams to triangulations of polygons, and then in subsection~\ref{sec: exact diagram matroidal props} we examine the matroidal properties of the exact subdiagrams.  Specifically, we show that a subdiagram of a Wilson loop diagrams $W$ defines a uniform matroid if and only if the subdiagram is exact (Theorem \ref{exactuniformthm}), and that this uniform matroid is equal to the dual of the matroid associated to $W$ restricted to the exact subdiagram (Remark \ref{remark exact dual restiction}).  Subsection~\ref{sec: matroids and equivalence} is devoted to the proof of Theorem~\ref{same matroid iff equiv}, along with an easy corollary which gives a formula for the size of each equivalence class (Corollary~\ref{number of equiv diagrams}).

Having identified the equivalence classes and their sizes, the next natural question is to identify the number of inequivalent diagrams under this equivalence relation.  The main result of Section~\ref{sec associahedron} is Theorem \ref{thm:count inequiv diagrams} which shows that the set of admissible Wilson loop diagrams (up to equivalence) is in bijection with the set of faces in the associahedron (up to parallelism).  Towards this result we go over some relevant background on polytopes in subsection~\ref{sec polytope background}, and then give the main result in subsection~\ref{sec associahedron results} in the form of two propositions: Propositions~\ref{prop easy way} and~\ref{prop hard way}.

@@

\begin{dfn}\label{equivdfn} 
Let $W = (\cP,[n])$ and $W' = (\cP',[n])$. Define a relation $\sim$ by $W\sim W'$ if
\begin{enumerate}
\item There exist exact subdiagrams $(P, V(P)) \subseteq W$ and $(P', V(P')) \subseteq W'$ with $V(P) =  V(P')$.
\item The complementary subdiagrams are identical, i.e. $(\cP \setminus P, V(P)^c) = (\cP' \setminus P', V(P')^c)$.
\end{enumerate}
The transitive closure of $\sim$ (which by an abuse of notation we also denote by $\sim$) defines an equivalence relation on the set of admissible Wilson loop diagrams with vertex set $[n]$.
\end{dfn}

@@

Throughout, we take the {\em matroid defined by $W$} to be the matroid $M(W)$ of Theorem \ref{thm WLD defines matroid}. Note from Definition~\ref{def:CWmatrix} that the vertices of the diagram $W$ correspond to columns of the associated matrix $C(W)$. Therefore, for any subset of vertices $V \subseteq [n]$, we can consider the restriction $C(W)|_V$ of $C(W)$ consisting of the columns corresponding to $V$. Since the non-zero entries of $C(W)$ are independent real variables, the matrix $C(W)|_V$ does not have full rank if and only if there are fewer than $|V|$ rows with non-zero entries. This is equivalent to saying that $V$ is a set of dependent columns of $C(W)$ if and only if $|\Prop(V)| < |V|$. Comparing this to Theorem \ref{thm:WLDmatroid}, we see that $M(W)$ is also the matroid realized by $C(W)$.

>>>>

This paper is the first in a two part series investigating the combinatorics and geometry underlying SYM $N=4$ theory. The series lays out several results about the relationship between Wilson loop diagrams and the positroid cells that give a CW-complex structure to the positive Grassmannian $\Gr(k,n)$. Understanding the precise nature of this relationship is crucial in the study of scattering amplitudes in SYM $N=4$, as described below.

In \cite{wilsonloop} the first author and Amat showed that every admissible Wilson loop diagrams can be identified with a unique positroid cell, and observed that this mapping is neither one-to-one nor onto. In this paper we give a simple characterization of this lack of injectivity in terms of an equivalence relation on the Wilson loop diagrams, and enumerate the number of different Wilson loop diagrams that correspond to each positroid cell. We also study the lack of surjectivity, and show that the number of distinct positroid cells associated to Wilson loop diagrams is given by counting the directions \hlfix{in}{of a?} associahedra.

The second paper \cite{generalcombinatoricsII} in this series identifies precisely which positroid cell is associated to a given Wilson loop diagram, by giving an algorithm for passing from a Wilson loop diagram to the Grassmann necklace of the associated positroid cell. In this manner the authors give a direct path from Wilson loop diagram to positroid cell that circumvents the painstaking process of examining bases sets. With this identification in hand, we also prove that any Wilson loop diagram with $k$ propagators corresponds to a $3k$ dimensional positroid cell, and characterize the volume forms that Wilson loop diagrams associate to each positroid cell.

In recent years, there has been an active program researching the geometry and combinatorics underlying SYM $N=4$ theory \cite{wilsonloop, Arkani-Hamed:2013jha, Amplituhedronsquared, galashinlam,AmplituhedronDecomposition}. This body of work started with the observation that the on shell (tree level) amplitudes of this theory correspond to the volume of a subspace of a positive Grassmannian, called an Amplituhedron \cite{Arkani-Hamed:2013jha}. Since then, there has been significant work studying the structure of the Amplituhedron both geometrically and combinatorially, for instance \cite{arkani:2012nw, Arkani-Hamed:2013kca, galashinlam,AmplituhedronDecomposition}. The focus of much of this work has been on understanding the relationship between BCFW diagrams and positroid cells, which are the combinatorial objects from the physics of the system and the corresponding natural objects in the positive Grassmannian respectively.

Meanwhile, there is a separate body of work studying SYM $N=4$ theory from the point of view of Feynman integrals in twistor space \cite{Adamo:2011pv, Britto:2005fq, Cachazo:2004kj}. These integrals are calculated via Wilson loops. As noted above, the \hlfix{first author and Amat}{At other points, we say "some of us" instead of specifying like this. Choose a convention} uncovered a connection between the Feynman integrals developed in this literature and positroids, defined  as matroids that are realized as an element of the positive Grassmannian $\Gr(k,n)$ \cite{wilsonloop}. In (separate) papers with the second author and Marcott, this work was extended to study these Feynman integrals in terms of positroid cells and the positive Grassmannians \cite{casestudy, non-orientable}. Others have tried different approaches to define a geometric space associated to these integrals in a manner similar to the Amplituhedron \cite{Amplituhedronsquared, HeslopStewart}.

Both the Amplituhedron literature and the Wilson loop literature associate $n$ particle $N^kMHV$ interactions in SYM $N=4$ theory to volume forms on a set of positroid cells of $\Gr(k,n)$. This gives a CW complex of some submanifold in $\Gr(k,n)$, and understanding the geometry and topology of these spaces is the subject of ongoing research.

However, the two approaches are not identical: for example, in the case of Wilson loop diagrams with $k=2$ and $n=6$, past work of some of us shows that the manifold is not contractible \cite{casestudy}. More generally, it has been shown that the spaces defined by the Amplituhedron and by Wilson loop diagrams are not the same \cite{Amplituhedronsquared}. In fact, it is in some ways more natural to consider Deodhar cells of $\mathbb{G}(k, n+1)$ for Wilson loop diagrams; in this case, one gets that the manifold represented by the diagrams is not orientable \cite{non-orientable}.

In this paper we focus on the correspondence between Wilson loop diagrams and positroids, which was first defined in \cite{wilsonloop}. Our main results are as follows. We first show that two Wilson loop diagrams define the same positroid if and only if they only differ by exact subdiagrams (Theorem~\ref{same matroid iff equiv}); equivalently, this says that the map from Wilson loop diagrams to positroid cells is one-to-one exactly on the subset of diagrams with no non-trivial exact subdiagrams. We count the total number of Wilson loop diagrams associated to a positroid cell (Corollary~\ref{number of equiv diagrams}). Finally, we show that enumerating the number of distinct positroid cells associated to Wilson loop diagrams is equivalent to enumerating the number of directions of an associahedron (Theorem~\ref{thm:count inequiv diagrams}).

%In that paper, \hlfix{one of us}{see previous comment about self reference convention} with Amat show that there are certain families of diagrams, specifically those with exact subdiagrams, that do map to the same positroid cell. In \cite{casestudy} some of us explicitly show that not every positroid cell of a given dimension corresponds to a Wilson loop diagram.

%In this paper, we characterize this correspondence more precisely. Specifically, we show that  We show that exact diagrams correspond to uniform matroids, and that contraction by exact subdiagrams has useful matroidal properties. Further, given a Wilson loop diagram with a non-trivial exact subdiagram, we count the total number of Wilson loop diagrams that correspond to that same positroid cell. While we do not count the number of distinct positroid cells that correspond to Wilson loop diagrams, we do show that this

@@

Section~\ref{section overall background} summarizes the required background, with definitions for Wilson loop diagrams given in subsection~\ref{section WLD background} and matroids in subsection~\ref{sec matroid background}. Key definitions include {\em admissible} Wilson loop diagrams (Definition~\ref{admisdfn}), {\em exact} subdiagrams (Definition~\ref{def:exact diagram}), and an equivalence relation on admissible Wilson loop diagrams (Definition~\ref{equivdfn}).  The connection between matroids and Wilson loop diagrams is stated precisely in Theorem \ref{thm WLD defines matroid}.

The main result of Section~\ref{sec equiv} is Theorem~\ref{same matroid iff equiv}, which says that two Wilson loop diagrams define the same matroid if and only if they are equivalent.  In subsection~\ref{sec: polygon partitions} we relate exact subdiagrams of Wilson loop diagrams to triangulations of polygons, and hence show that every Wilson loop diagram can be uniquely decomposed as a collection of exact subdiagrams (Corollary~\ref{uniqueproppartitioncor}). In subsection~\ref{sec: exact diagram matroidal props} we examine the matroidal properties of the exact subdiagrams, and prove that a subdiagram of a Wilson loop diagrams defines a uniform matroid if and only if the subdiagram is exact (Theorem \ref{exactuniformthm}). This uniform matroid is in fact equal to the dual matroid restricted to the exact subdiagram (Remark \ref{remark exact dual restiction}).  Finally, subsection~\ref{sec: matroids and equivalence} is devoted to the proof of Theorem~\ref{same matroid iff equiv}, along with an easy corollary which gives a formula for the size of each equivalence class (Corollary~\ref{number of equiv diagrams}).

Having identified the equivalence classes and their sizes, the next natural question is to identify the number of inequivalent diagrams under this equivalence relation.  The main result of Section~\ref{sec associahedron} is Theorem \ref{thm:count inequiv diagrams}, which shows that the set of admissible Wilson loop diagrams (up to equivalence) is in bijection with the set of faces in the associahedron (up to parallelism).  Towards this result we go over some relevant background on polytopes in subsection~\ref{sec polytope background}, and then prove the main result in subsection~\ref{sec associahedron results} in the form of two propositions: Proposition~\ref{prop easy way} and Proposition~\ref{prop hard way}.

@@

\begin{dfn}\label{equivdfn}

Let $W = (\cP,[n])$ and $W' = (\cP',[n])$, and say that $W$ and $W'$ are related if
\begin{enumerate}
\item There exist exact subdiagrams $(P, V(P)) \subseteq W$ and $(P', V(P')) \subseteq W'$ with $V(P) =  V(P')$.
\item The complementary subdiagrams are identical, i.e. $(\cP \setminus P, V(P)^c) = (\cP' \setminus P', V(P')^c)$.
\end{enumerate}
Let $\sim$ denote the transitive closure of this relation. This is an equivalence relation on the set of all admissible Wilson loop diagrams with vertex set $[n]$.
\end{dfn}

@@

Throughout, we take the {\em matroid defined by $W$} to be the matroid $M(W)$ of Theorem \ref{thm WLD defines matroid}. Note from Definition~\ref{def:CWmatrix} that the vertices of the diagram $W$ correspond to columns of the associated matrix $C(W)$. Therefore, for any subset of vertices $V \subseteq [n]$, we can consider the restriction $C(W)|_V$ of $C(W)$ consisting of the columns corresponding to $V$. Since the non-zero entries of $C(W)$ are independent real variables, \hlfix{the matrix $C(W)|_V$ does not have full rank if and only if there are fewer than $|V|$ rows with non-zero entries.}{I don't think this is quite true?} This is equivalent to saying that $V$ is a set of dependent columns of $C(W)$ if and only if $|\Prop(V)| < |V|$. Comparing this to Theorem \ref{thm:WLDmatroid}, we see that $M(W)$ is also the matroid realized by $C(W)$.
